## R packages to install {.unnumbered}

Here is a list of packages that you need to install so that you can follow the practicals. You can use `install.packages(name-package)`. 

### For day 2 
```{r, eval=FALSE}

library(bipartite)
library(igraph)
library(vegan)
library(cheddar)
library(matlib)
library(sbm)
library(alluvial)
library(econetwork)
library(poweRlaw)

```



Please source or copy-paste in your R session the functions from the `crutch_functions.R` file (available on the [data repository of the course](https://github.com/econetoolbox/data/) in the day2_files folder).


In the course script, there is also an alternative to get the database of all the food webs that are in this package, but it requires to have `rmangal` installed:


```{r, eval=FALSE}

library(rmangal)
mg1 <- create.fw.list(db="mg", ref=TRUE, spatial = TRUE) # this function can be found in the 'crutch_functions.R' script.

```



### For days 3 and 4

```{r, eval=FALSE}

remotes::install_github("Jo-Theo/shinySbm")
remotes::install_github("Chabert-Liddell/colSBM")
library(colSBM)
library(network)
library(GGally)
library(RColorBrewer)
library(knitr)

```









