## R packages to install {.unnumbered}

Here is a list of packages that you need to install so that you can follow the practicals. You can use `install.packages(name-package)`. 

### For day 2 
```{r, eval=FALSE}

library(bipartite)
library(igraph)
library(FWebs)
library(vegan)
library(cheddar)
library(matlib)
library(sbm)
library(alluvial)
library(econetwork)
library(poweRlaw)

```



Mac users won't be able to install the package `FWebs`. They can use the file `crutch_functions.R` (available on the [data repository of the course](https://github.com/econetoolbox/data/) in the day2_files folder), which contains the needed functions. 


In the course script, there is also an alternative to get the database of all the food webs that are in this package, but it requires to have `rmangal` installed:


```{r, eval=FALSE}

library(rmangal)
mg1 <- create.fw.list(db="mg", ref=TRUE, spatial = TRUE)

```



### For days 3 and 4

```{r, eval=FALSE}

remotes::install_github("Jo-Theo/shinySbm")
remotes::install_github("Chabert-Liddell/colSBM")
library(colSBM)
library(network)
library(GGally)
library(RColorBrewer)
library(knitr)

```









